\Huge \textbf{Species sympatry shapes brain size evolution in Primates}

### 2021-08-19 Article/Article.Rmd
Update following benoit's correction + putting right frugivory threshold + filtering frugivorous correctly + adding sympatry vs diversification

header-includes:
# - \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  #- \usepackage{lineno} # For line numbering
  #- \linenumbers # For line numbering
  #- \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  - \pagenumbering{gobble} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
      \setcounter{table}{0}  
      \renewcommand{\thetable}{S\arabic{table}} 
      \setcounter{figure}{0} 
      \renewcommand{\thefigure}{S\arabic{figure}}
    }
---

@@

load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversification_withautocorr.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdirectionSelection.RData")

\captionsetup{list=no}

<!-- The front page -->
<!-- \centering -->
<!-- \raggedright -->
<!-- \newpage -->
<!-- \tableofcontents -->
\newpage

@@

**Abstract |** In hypotheses related to the evolution of animal intelligence, a particular emphasis was given to the role of intra-species conspecifics. Yet, one species rarely occupies an area alone: Space is a place shared between many species. These latter can be new competitors to outperform or cooperators to decode and maneuver, thereby stimulating cognition. They can also contribute to modifying and depleting the environment such as it overly increases its complexity, thereby alleviating the benefice of cognition. Considering brain size as a proxy for cognition, we used primates as a study example to test for the intertwine between species co-occurrence and cognition. We retraced the evolutionary history of several brain areas considering competitive or non-competitive evolutionary scenarios. We found that competitive evolutionary scenarios best predicted the evolution of some brain areas that were related to interacting with the social or ecological environment, with a decrease of their relative size the higher the sympatry rate. The whole brain size, as well as the size of areas used in more general contexts were not related to the sympatry rate. In any case, the evolutionary success of primate species was unrelated th the relative sizes of brain areas. Overall, this comparative study suggests that species co-occurrence contributes to shaping the selective pressure acting upon animal cognition, yet leaves open the question of the underlying ecological mechanisms at play.

@@

\faTags \hspace{0.01in} **Keywords:** Brain size - Cognition - Competition - Diversification - Intelligence evolution - Primates - Species co-occurrence

<!-- TC:ignore -->
\faInfoCircle  \hspace{0.01in} **Word Count (main text only):** `r words$num_words` \newline
<!-- Character Count: `r words$num_char` -->
<!-- TC:endignore -->

@@

On the road to brain size evolution, generally considered as an equivalent of cognition evolution, mysteries are plenty [@van2006some; @dunbar2017there]. Overall, the evolution of the brain is the consequence of constraints (e.g. energetic limitations, @navarrete2011energetics), but also of socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference]. Primates are pivotal species to study the evolution of cognition given the complexity of their socio-ecological environment and the inevitable implication for retracing human evolutionary history [@byrne2000evolution]. Their cognition can result from interactions of individuals with conspecifics within the social unit (*Social Intelligence Hypothesis*, @byrne1994machiavellian; @dunbar2017there), among generations (*Cultural Intelligence Hypothesis*, @whiten2007evolution; @reader2002social, also known as the *Vygotskian Intelligence Hypothesis* in humans, @herrmann2007humans; @tomasello2009cultural; @wilson1991molecular; @van2011social), between social units (*Napoleonic Intelligence Hypothesis*, @ashton2020interactions), or with the rest of their environment (*Ecological Intelligence Hypothesis*; @clutton1980primates; @milton1981distribution; @rosati2017foraging).

|   A particular emphasis was given to the role of intra-species conspecifics in shaping animals cognitive abilities. Yet, space is a place shared between a plethora of species that can interact directly or indirectly, in particular if sharing some ecological traits, especially diet [@drury2018contrasting]. To settle to new coral reefs, fishes use pops and clicks of other fishes as an honest signal for resourcefulness there [@gordon2019acoustic], mangabeys follow calls from hornbills to locate fruiting trees [@olupot1998fruit], and interactions even happen across kingdoms, with migratory birds interpreting phenological cues as synonymous of insect availability [@mcgrath2009flower].

|   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities (but see discussion in @logan2018beyond). Although generally considered as a singular entity, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic]. For instance, the Striatum is stimulated during social interaction [@baez2013role]. Therefore, its size is expected to be positively influenced when contacts with other species increase, for instance in species performing mixed-species groups (e.g. in callhitrichine primates, @heymann2015unlike). Conversely, the evolution of the Main Olfactory Bulb (MOB), allowing processing olfactory cues, or the Hippocampus, home of a spatio-temporal memory [@burgess2002human], should be linked to environmental complexity faced by a species, in particular when foraging. This environmental complexity can increase if space is shared with species with similar dietary habits because it contributes to impoverishing the environment and increasing its unpredictability, thus looping on selective pressure towards cognitive abilities [@grove2013evolution: @robirainreview]. Given high energetic costs of the brain [@raichle2006brain], it is thus expected that sympatry should induce positive selection on the size of brain areas involved in foraging-related information processing, unless the induced environmental complexity starts being too high for cognitive foraging to be adaptive. Given that the Striatum also underpins reward expectation and action, goal-directed behaviour and planning abilities [@johnson2007integrating], the same scenario might apply as well to the Striatum of species that do not live in mixed-groups, which actually constitute the majority of the primate lineage. In between, more generalist areas, such as the Cerebellum [@koziol2014consensus; @sokolov2017cerebellum] and the Neocortex [@wiltgen2004new] underlying movement and/or general information processing and retention, could also be stimulated by the presence of other species movement traces, noises or odors to decode.

|    Recent tools have been developed to infer the effect of species interactions on trait evolution, either by modelling trait divergence in co-occurring species from a same guild (e.g. dietary guild; Matching Competition: MC models) or considering that the evolutionary rate depends on the number of co-occurring lineages within the guild (density dependence; linear: DD$_{lin}$ or exponential: DD$_{exp}$; @drury2016estimating). After reconstructing primate biogeography history (N$_{species}$ = `r minRange`, @matzke2013probabilistic; @matzke2016stochastic) when considering `r length(areaName)` biogeographic areas (Figure \@ref(fig:figmap); @kamilar2009environmental) and diet evolution (N$_species$ = `r minFruit + minLeaf` to `r maxFruit + maxLeaf`; @bollback2006simmap), we calculated the likelihoods of models considering the role of species interactions (including competitive scenarios) in the evolution of either the whole brain (using the encephalic quotient, EQ, as a proxy, N$_{species}$ = `r minEQ` to `r maxEQ`), or the relative size of specific brain areas associated with foraging-related information perception, processing or retention (Figure. \@ref(fig:figbrain); N$_{species}$ = `r minAllAreas` to `r max(maxNeocortex,maxHippocampus,maxCerebellum,maxStriatum,maxMOB)`). The use of specific region size relatively to the body mass and not raw size depicts the evolution of cognitive abilities in terms of allocation rather than abilities per se (but see discussion in @deaner2000comparative). We also estimated the likelihoods of simpler models assuming no effect of species interactions, like the Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) considering that traits are constrained around on optimal value (e.g. stabilizing selection; see @blomberg2020beyond for a review) or the Early-Burst model (EB, @blomberg2003testing), this latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density effect is not an artefact due to time-dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weights of the Akaike Information Criterions corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see [Models of trait evolution: does interspecific interactions shape brain size evolution?]).

@@

|   Non-competitive models were the most likely in describing the evolutionary history of the EQ, the Neocortex and the Cerebellum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), two areas specifically involved in movement and/or general information processing [@wiltgen2004new; @koziol2014consensus; @sokolov2017cerebellum] but also in memory consolidation for the Neocortex [@wiltgen2004new]. By contrast, competitive models were most supported in areas involved in short-term working memory and long-term spatio-temporal information retention (Hippocampus, @burgess2002human) and in planning, reward assessment, and information processing during social interaction (Striatum, @baez2013role; @johnson2007integrating), while they tended to be so for one involved in sensory abilities (the main olfactory bulb, MOB; Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)). When density-dependent models were the best fit, the rate (*r*, Figure \@ref(fig:figresultsevolution)) suggested an acceleration of the evolutionary tempo together with increased lineage diversity for the Hippocampus and the Striatum, but a slowdown for the MOB.

@@

|   Next, to understand the directionality of the selection gradient shaped by co-occurrence (i.e. selection for “bigger” or “smaller” brain if the more species), we fitted phylogenetic regressions (see [Phylogenetic regressions] a)). For these linear regressions, the predicted variable was the relative brain size values of the different areas. We considered the average surface of the frugivorous species range that was overlapped by other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire distribution range as covariates. On average ($\pm$ SE), the considered primate species had `r meanOverlap*100`% of their range overlapping with other species ($\pm$ `r seOverlap*100`). That ranged from `r minOverlapSpecies*100`% of overlap (*`r whichSpeciesMinOverlap`*), to `r maxOverlapSpecies*100`% of overlap (*`r whichSpeciesMaxOverlap`*). In term of distribution range, the considered primate species co-occurred on average with `r meanNspeciesCoocc` other primate species ($\pm$ `r seNspeciesCoocc`), ranging from `r minNspeciesCoocc` other species (*`r whichSpeciesMinNCoocc`*), to `r maxNspeciesCoocc` species (*`r whichSpeciesMaxNCoocc`*). The number of sympatric species never influenced significantly the relative size of the brain or other specific areas  (Table \@ref(tab:tableRegGradient)). Conversely, we found that the percentage of range shared with other species correlated with the relative size of areas for which the evolutionary history was better described with competitive models: The Hippocampus, the MOB and the Striatum (Hippocampus: $t$`r testGradientHippocampus`;  MOB: $t$`r testGradientMOB`; Striatum: $t$`r testGradientStriatum`). The correlations were all negative (Hippocampus: `r estimateGradientHippocampus`;  MOB: `r estimateGradientMOB`; Striatum: `r estimateGradientStriatum`), which means that higher species overlap associates with lower relative size, insensitive to data and phylogeny variability (Table \@ref(tab:tabledfsensitivity)). Thus, it suggests that sympatric species are either subject to a positive selection for smaller brain, or to a less stringent selection on advanced cognitive abilities.

@@

|   Finally, we investigated the evolutionary consequences of brain evolution by evaluating whether brain sizes correlated with species diversification rates (i.e. speciation minus extinction rates), by using lineage-specific birth-death models of species diversification [@maliet2019model; @maliet2020fast]. Overall, species diversification, estimated based on molecular phylogeny increased over time (Figure \@ref(fig:figdiversificationTime)), particularly in the early and late Miocene, around `r dateFirstRupt[2]` (CI95% = `r paste("[",dateFirstRupt[3],",",dateFirstRupt[1],"]", sep="")`) and `r dateSecondRupt[2]` (CI95% = `r paste("[",dateSecondRupt[3],",",dateSecondRupt[1],"]", sep="")`) Myr ago (Figure \@ref(fig:figdiversificationTime)). When accounting for phylogenetic dependence, no significant relationship between diversification rate and the relative size of brain areas was found (Table \@ref(tab:tableRegDiversification); see robustness in Table \@ref(tab:tabledfsensitivity2)).

@@

```{r calculationValueDiscussion}
#Import data
summaryDataForPlot <- read_delim("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/OutputEvolModel/Dataplot.txt","\t", escape_double = FALSE, trim_ws = TRUE)

The bigger is not necessarily the better. The size of the brain is subject to a compromise between the energy it incurs, and the increase of fitness it allows. This is clearly emphasized by the fact that the biggest areas, the Cerebellum and the Neocortex, as well as the whole brain (EQ) were best described by the Orstein-Uhlenbeck process, what might suggest a stabilisation towards an optimal size as a result of an equilibrium between costs and benefices. As the brain area is regionally specialised [@barton2000mosaic], it was also expected that brain regions could be under different selective pressures. For instance, the Lemuriformes, that are known to prioritize smell compared to other primate species, indeed have the largest relative MOB size (i.e. pondered by body size) in our data (Lemuriformes: mean $\pm$ SE = `r meanseMOBlemu[1]` $\pm$ `r meanseMOBlemu[2]`, other: `r meanseMOBother[1]` $\pm$ `r meanseMOBother[2]`, \@ref(fig:figbrain)). Similarly, Platyrrhini, and Callitrichine in particular, are known to form poly-specific associations [@heymann2000behavioural] and indeed show the highest relative size of the Striatum in our data (Platyrrhini: mean $\pm$ SE = `r meanseStriatumplaty[1]` $\pm$ `r meanseStriatumplaty[2]`, other: `r meanseStriatumother[1]` $\pm$ `r meanseStriatumother[2]`, \@ref(fig:figbrain)). Here, we show that sympatry is one factor that affects the selective regime under which each brain region evolves: Although the brain as a whole was insensitive to species co-occurrence, this latter nonetheless induced a change in the relative size of the Hippocampus, and the Striatum, and potentially in the MOB. These areas imply individual-based and social-based information processing, pinpointing that the two components might be under selection in primates. This supports the general discussion on the importance of social *vs* ecological factors to explain primate cognition evolution and diversity [@decasien2017primate; @powell2017re].

|   The relative size of these brain areas were negatively associated with one index of sympatry intensity, the average overlap of a species distribution range with that of other species. Because the ratio of the other brain areas is insensitive to sympatry intensity, it is unlikely that this results from a by-product of a selection on body mass and might thus be a consequence of species direct or indirect interactions. In general, competition is the first-thought mechanism to describe community structures [e.g. @rocha2015role]. In primates, it is rare to observe direct competition between species, which should translate into site exclusion. In fact, inter-species site exclusion in this taxa has only been observed in gibbons [@suwanvecho2012interspecific]. For species with overlapping diet, indirect competition for resource might happen. It has been shown that individuals tend to use social or environmental cues depending on their reliability [@rafacz2003environmental; @dunlap2016foraging]. The utilization of these cues should promote cognition until reaching a certain range of environmental complexity [@grove2013evolution; @robirainreview]. Our results could thus suggest that species sympatry may trigger an important decrease in cues reliability, for instance due to higher and unforeseen depletion rate. That would explain a relaxed positive selection for advance cognitive abilities, and conversely, given the cost of brain maintenance [@raichle2006brain], would induce a negative selection towards areas involved in foraging related activities like the Hippocampus, the MOB and the Striatum.

>>>>

header-includes:
  - \usepackage{lettrine}
# - \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  #- \usepackage{lineno} # For line numbering
  #- \linenumbers # For line numbering
  #- \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  - \pagenumbering{gobble} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
      \setcounter{table}{0}  
      \renewcommand{\thetable}{S\arabic{table}} 
      \setcounter{figure}{0} 
      \renewcommand{\thefigure}{S\arabic{figure}}
    }
---

<!-- ------------------------------------------------------------------- -->
<!-- ---------------------      LAYOUT       --------------------------- -->
<!-- ------------------------------------------------------------------- -->

<!-- TO PUT THE LETTRINE -->

\newcommand{\initial}[1]{%
	\lettrine[lraise=0, loversize=0.5,nindent=0em]{
		\color{black}
     		{\textsc{#1}}}{}}

<!-- TO CREATE GRAY BOX -->

<!-- ------------------------------------------------------------------- -->
<!-- ------------------------------------------------------------------- -->
<!-- ------------------------------------------------------------------- -->

@@

load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversification_withautocorr.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdirectionSelection.RData")
load("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/REnvironments/PGLSdiversificationAndSympatry.RData")
```

\captionsetup{list=no}
\newpage

@@

**Abstract |** Hypotheses related to animal intelligence evolution highlight the role of intra-species conspecifics. Yet, one species rarely occupies an area alone: Space is a place often occupied by many species, sometimes belonging to the same ecological guild. These species can act as competitors to outperform or cooperators to decode and maneuver, thereby stimulating cognition. They can also contribute to modifying and depleting the environment such as it overly increases depletion intensity and makes it poorly predictable, thereby alleviating the benefice of cognition. Considering brain size as a proxy for cognition, we used primates as a study example to test for the intertwine between species sympatry and cognition. We retraced the evolutionary history of several brain areas considering competitive or non-competitive evolutionary scenarios. We found that competitive evolutionary scenarios best predicted the evolution of brain areas related to interacting with the social or ecological environment, with a decrease of their relative size the higher the sympatry rate. The whole brain size, as well as the size of areas used in more general contexts were not related to the sympatry rate. In any case, the evolutionary success of primate species was unrelated to the relative sizes of brain areas but was affected by sympatry. Overall, this comparative study suggests that species sympatry, which likely convolutes the environment, contributes to shaping the selective pressure acting upon primate cognition and diversification.

@@

\faTags \hspace{0.01in} **Keywords:** Brain size - Cognition - Competition - Co-occurrence - Diversification - Intelligence evolution - Primates - Sympatry

\faInfoCircle  \hspace{0.01in} **Word Count:** `r words$num_words` \newline
<!-- Character Count: `r words$num_char` -->

@@

<!-- On the road to brain size evolution, generally considered as an equivalent of cognition evolution, mysteries are plenty [@van2006some; @dunbar2017there]. Overall, the evolution of the brain is the consequence of constraints (e.g. energetic limitations, @navarrete2011energetics), but also of socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference].  -->
<!-- Primates are pivotal species to study the evolution of cognition given the complexity of their socio-ecological environment and the inevitable implication for retracing human evolutionary history [@byrne2000evolution].  -->

\initial{\textcolor{black}{C}}ognition evolution results from the balance between socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference] and physiological and energetic constraints [@navarrete2011energetics]. Primates are pivotal species for cognitive studies [@byrne2000evolution] because their cognition is thought to be promoted by interactions of individuals with conspecifics within the social unit [@byrne1994machiavellian; @dunbar2017there], among generations [@wilson1991molecular; @whiten2007evolution; @reader2002social; @herrmann2007humans; @tomasello2009cultural; @van2011social], between social units [@ashton2020interactions], or with the rest of their environment [@clutton1980primates; @milton1981distribution; @rosati2017foraging].

|   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities (but see discussion in @logan2018beyond). Although generally considered as a singular entity, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic]. For instance, the Striatum is stimulated during social interactions [@baez2013role]. Its size is thus expected to be positively influenced when contacts with other species increase, as in species performing mixed-species groups, a rare  fact in primates nonetheless (e.g. in callitrichine primates, @heymann2015unlike). The Striatum also underpins reward expectation and action, goal-directed behaviour and planning abilities [@johnson2007integrating] which is key when foraging. Foraging-related areas, as are also the Main Olfactory Bulb (MOB) or the Hippocampus, home of a spatio-temporal memory [@burgess2002human], are expected to be influenced by the ecological environment. Fruits being the archetype of a hard-to-find resource yet predictable [@janmaat2016spatio; @robirabotany], this might be particularly relevant for frugivorous primates. The changes induced by other species with similar dietary habits can contribute to increasing its unpredictability, thus changing the selective pressure applying to cognitive abilities [@grove2013evolution: @robirainreview]. Due to the cost of maintaining the brain functional [@raichle2006brain], it expected that the increase in environmental unpredictable only stimulates cognition up to a certain threshold (i;e. positive selection for bigger brain), before it is such high that smaller brain sizes are more adaptive. In between, more generalist areas, such as the Cerebellum [@koziol2014consensus; @sokolov2017cerebellum] and the Neocortex [@wiltgen2004new] underlying movement and/or general information processing and retention, could also be stimulated by the presence of other species movement traces, noises or odors to decode.

|    Here, we investigated the intertwine between species sympatry and cognition using frugivorous primate as study example. To infer the effect of species interactions on brain size evolution within the frugivorous primates, we evaluated the support for competitive or non-competitive evolutionary scenarios in the context of sympatry and investigated the directionality of the selection upon brain size evolution in the context of sympatry. Finally, we tested for correlative pattern of the evolutionary success in primates with brain size and current sympatry context.

@@

```{r calculationValueForFirstResults}
#Import data
summaryDataForPlot <- read_delim("C:/Users/robira/Documents/PhD/Meta_analysis/Meta_analysis_cognition_primates/OutputEvolModel/Dataplot.txt","\t", escape_double = FALSE, trim_ws = TRUE)

We gathered a large database on the whole brain size, the size of the Cerebellum and the Neocortex, involved in daily information processing and retention, and the Main Olfactory Bulb (MOB), the Striatum, and the Hippocampus, those three latter being related to social and foraging-related tasks, containing between `r minAllAreas` to `r maxEQ` frugivorous primate species (depending on the brain area considered). It evidenced ample variations in brain size. For instance, the lemuriformes, that are known to prioritize smell compared to other primate species, have the largest relative MOB size (i.e. pondered by whole-body mass) in our data (Lemuriformes: mean $\pm$ SE = `r meanseMOBlemu[1]` $\pm$ `r meanseMOBlemu[2]`, other: `r meanseMOBother[1]` $\pm$ `r meanseMOBother[2]`, \@ref(fig:figbrain)). Similarly, platyrrhini, and callitrichine primates in particular, are known to form poly-specific associations [@heymann2000behavioural] and indeed show the highest relative size of the Striatum in our data (Platyrrhini: mean $\pm$ SE = `r meanseStriatumplaty[1]` $\pm$ `r meanseStriatumplaty[2]`, other: `r meanseStriatumother[1]` $\pm$ `r meanseStriatumother[2]`, \@ref(fig:figbrain)).

|   We first reconstructed primate biogeography history (N$_{species}$ = `r minRange`, @matzke2013probabilistic; @matzke2016stochastic) when considering `r length(areaName)` biogeographic areas (Figure \@ref(fig:figmap); @kamilar2009environmental) and their diet evolution (N$_{species}$ = `r minFruit + minLeaf` to `r maxFruit + maxLeaf`; @bollback2006simmap). We then calculated the likelihoods of models considering the role of species interactions (matching competition MC models, @nuismer2015predicting; models of linear or exponential density dependence, DD$_{lin}$and  DD$_{exp}$, of the evolutionary rate on the number of sympatric lineages; @drury2016estimating see [Models of trait evolution: does interspecific interactions shape brain size evolution?]) in the evolution of either the whole brain (using the encephalic quotient, EQ, as a proxy), or the size of the aforementioned specific brain areas relative to the whole-body mass (Figure. \@ref(fig:figbrain)) and compared their likelihood to those of simpler models assuming no effect of species interactions, like the Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) considering that traits are constrained around on optimal value (e.g. stabilizing selection; see @blomberg2020beyond for a review) or the Early-Burst model (EB, @blomberg2003testing), this latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density effect is not an artefact due to time-dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weights of the Akaike Information Criterions corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see [Models of trait evolution: does interspecific interactions shape brain size evolution?]).

|   We found that non-competitive models were the most likely in describing the evolutionary history of the EQ, the Neocortex and the Cerebellum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), two areas specifically involved in movement and/or general information processing [@wiltgen2004new; @koziol2014consensus; @sokolov2017cerebellum] but also in memory consolidation for the Neocortex [@wiltgen2004new]. The fact that these biggest areas are best described by the Ornstein-Uhlenbeck process suggests a stabilization towards an optimal size and illustrates the trade-offs between costs and benefices of brain development [@isler2009expensive]. By contrast, competitive models were most supported in areas involved in foraging-related and social areas, either specialised in spatio-temporal memory (Hippocampus, @burgess2002human) or planning abilities and reward assessment (Striatum, @baez2013role; @johnson2007integrating), while they tended to be so for one involved in sensory abilities (the main olfactory bulb, MOB; Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)). When density-dependent models were the best fit, the rate (*r*, Figure \@ref(fig:figresultsevolution)) suggested an acceleration of the evolutionary tempo together with increased lineage diversity for the Hippocampus and the Striatum, but a slowdown for the MOB.

@@

|   Next, to understand the directionality of the selection gradient shaped by sympatry (i.e. selection for “bigger” or “smaller” brain the more species), we fitted phylogenetic regressions (see [Phylogenetic regressions] a)). For these linear regressions, the predicted variable was the relative brain size values of the different areas. We considered the average surface of the frugivorous species range that was overlapped by other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire distribution range as covariates based on IUCN date [@IUCN]. On average ($\pm$ SE), the considered primate species had `r meanOverlap*100`% of their range overlapping with other species ($\pm$ `r seOverlap*100`). That ranged from `r minOverlapSpecies*100`% of overlap (*`r whichSpeciesMinOverlap`*), to `r maxOverlapSpecies*100`% of overlap (*`r whichSpeciesMaxOverlap`*). In terms of distribution range, the considered primate species co-occurred on average with `r meanNspeciesCoocc` other primate species ($\pm$ `r seNspeciesCoocc`), ranging from `r minNspeciesCoocc` other species (*`r whichSpeciesMinNCoocc`*), to `r maxNspeciesCoocc` species (*`r whichSpeciesMaxNCoocc`*). The number of sympatric species never influenced significantly the relative size of the brain or other specific areas (Table \@ref(tab:tableRegGradient)). Conversely, we found that the percentage of range shared with other species correlated with the relative size of areas that were better fit with competitive models: The Hippocampus, and the Striatum (Hippocampus: $t$`r testGradientHippocampus`; Striatum: $t$`r testGradientStriatum`). The correlations were all negative (Hippocampus: `r estimateGradientHippocampus`; Striatum: `r estimateGradientStriatum`), which means that higher species overlap associates with lower relative size, insensitive to data and phylogenetic uncertainty (Table \@ref(tab:tabledfsensitivity), Figure \@ref(fig:forestPlot)). Thus, it suggests that sympatric species are either subject to a positive selection towards smaller brain, or to a less intense selection for advanced cognitive abilities compared with isolated species.

@@

|   In addition, we investigated the evolutionary consequences of brain evolution by evaluating whether brain sizes correlated with species diversification rates (i.e. speciation minus extinction rates), by using lineage-specific birth-death models of species diversification [@maliet2019model; @maliet2020fast]. Overall, species diversification, estimated based on molecular phylogeny increased over time (Figure \@ref(fig:figdiversificationTime)), particularly in the early and late Miocene, around `r dateFirstRupt[2]` (CI95% = `r paste("[",dateFirstRupt[3],",",dateFirstRupt[1],"]", sep="")`) and `r dateSecondRupt[2]` (CI95% = `r paste("[",dateSecondRupt[3],",",dateSecondRupt[1],"]", sep="")`) Myr ago (Figure \@ref(fig:figdiversificationTime)). When accounting for phylogenetic dependence, no significant relationship between diversification rate and the relative size of brain areas was found (Table \@ref(tab:tableRegDiversification), Figure \@ref(fig:forestPlot); see robustness in Table \@ref(tab:tabledfsensitivity2)). Given the context dependence of the direction of selection (towards bigger sizes in the absence of other species, smaller sizes otherwise), there is no surprise that we do not observe a correlation between evolutionary success (approximated by diversification rate) and the three studied areas affected by inter-species sympatry. Surprisingly however, we found no positive association between the EQ and the diversification rate, nor of that latter with the relative size of the Cerebellum or the Neocortex, two areas insensitive to inter-species sympatry too. This is puzzling because this contradicts a recent study [@melchionna2020macroevolutionary]. The visual inspection of the regressions however clearly evidenced a positive trend for the EQ and the Neocortex if discarding phylogenetic signals (Figure \@ref(fig:figRegressionDiversification)). In fact, a sudden encephalisation in primates is clearly associated to a limited number of close-related species [@decasien2017primate; @melchionna2020macroevolutionary]. This clearly limits the statistical power of our phylogenetically corrected analyses as we cannot decipher whether the connivance between brain size and evolutionary success results from a true biological link or appeared by chance. A positive association between brain size and diversification was also found in birds [@sayol2019larger] given that bigger brains act as a buffer to environmental challenge, thus boosting individuals' fitness [@sol2007big]. This means that, despite what we found here, the positive association between brain size and evolutionary success evidenced in @melchionna2020macroevolutionary remains a likely possibility. Finally, despite diversification was uncorrelated with brain size in frugivorous primates, it was influenced by the sympatry context. In particular, phylogenetic regressions highlighted a negative effect of the number of sympatric species on the diversification rate (`r textEstOneVar(as.numcharac(results.df_diversificationAndSympatry[4,2:4]))`, `r textTestOneVar(as.numcharac(results.df_diversificationAndSympatry[4,5:6]), statistics="t")`, Table \@ref(tab:tableRegDiversificationAndSympatry), Figure \@ref(fig:forestPlot)). In other words, the higher the number of sympatric species, the lower the diversification, an inverted pattern compared to what is found in New World passerines [@kennedy2017historical].

@@

The bigger is not necessarily the better. The size of the brain is subject to a compromise between the energy it incurs, and the increase of fitness it allows. This is clearly emphasized by the fact that the biggest areas, the Cerebellum and the Neocortex, as well as the whole brain (EQ) were best described by the Orstein-Uhlenbeck process, what might suggest a stabilisation towards an optimal size as a result of an equilibrium between costs and benefices. As the brain area is regionally specialised [@barton2000mosaic], brain regions could be under different selective pressures as suggested by the differences we highlighted in some brain areas (e.g. MOB or Striatum) across different primate genus. Here, we further show that sympatry is one factor that affects the selective regime under which each brain region evolves: Although the brain as a whole was insensitive to species sympatry, this latter nonetheless induced a change in the relative size of the Hippocampus, and the Striatum, and potentially in the MOB. These areas imply individual-based and social-based information processing, pinpointing that the two components might be under selection in primates [@decasien2017primate; @powell2017re; @gonzalez2018inference]. Overall, the relative size of foraging-related brain areas were negatively associated with one index of sympatry intensity (the range overlap with other species) and might thus be a consequence of species direct or indirect interactions. In general, competition is the first-thought mechanism to describe community structures [e.g. @rocha2015role]. In primates, it is rare to observe direct competition between species, which should translate into site exclusion. In fact, inter-species site exclusion in this taxa has only been observed in gibbons [@suwanvecho2012interspecific]. For species with overlapping diet, indirect competition for resource might happen. It has been shown that individuals tend to use social or environmental cues depending on their reliability [@rafacz2003environmental; @dunlap2016foraging]. The utilization of these cues should promote cognition until reaching a certain range of environmental complexity [@grove2013evolution; @robirainreview]. Our results could thus suggest that species sympatry may trigger an important decrease in cues reliability, for instance due to higher and unforeseen depletion rate. That would explain a relaxed positive selection for advance cognitive abilities, and conversely, given the cost of brain maintenance [@raichle2006brain], would induce a negative selection towards areas involved in foraging related activities like the Hippocampus, the MOB and the Striatum. Not only was brain evolution affected by sympatry, but sympatry (the maximum number of co-occurring species only) induced a slowdown in diversification, which contrasts with the picture offered by passerine birds at finer resolution [@kennedy2017historical]. Density-dependence within clade is a known factor to affect negatively diversification rate [@moen2014does]. In particular, species competing for resource are thought to contribute to limiting competitors' range [@price2009evolutionarily], hence to constrain population size and diversification rate [@pigot2013species]. Given the large resolution and low overlap threshold to consider sympatry considered here (i.e. overlap of `r geographicThresholdVector[1]*100`% of the distribution area), this competitive view fits the observed influence of sympatry on frugivorous primate diversification.

|   The use of brain size as a proxy for cognition is a central debate with no optimal solution (see grounded criticism from @deaner2000comparative; @healy2007critique; @logan2018beyond). The current flourishment of consortia, allowing for much more detailed and standardized anatomical measurements (e.g. in primates: @milham2018open), or with standardized behaviourally explicit comparisons (e.g. on captive [@many2019establishing] or wild [@janmaat2021using] primates), might alleviate biases stemming from brain size analysis, but this will take time to generate large-enough datasets. In the meanwhile, brain size is a proxy much appreciated in practice, because of its easy accessibility for a “large” number of species. Here, we showed that species sympatry can influence the evolutionary scenario that applies to animals' brain, but the proximate mechanisms at play remain to be elucidated. Finally, it is very likely that any hypothesis on cognition evolution, generally discussed within species, could be broadened to a between-species context: Polyspecific social associations do exist [@porter2001benefits], as well as inter-species territory defense [@drury2020competition; @losin2016ecological] or imitation and copying [@persson2018spontaneous; @pepperberg2002allospecific]. As Alice said “It's a great huge game of chess that's being played—all over the world” (@Carroll, Chapter II) and all individuals are just pieces to play with or against, no matter the species.
